---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Nargis Taraki"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 Set up working directory and load necessary libraries
library(tidyverse)
library(agricolae) 
library(lubridate)
library(here)
library(ggplot2)

here()

# Import raw data and set date columns
NTL_LTER_Data <- read.csv(here('Data','Raw','NTL-LTER_Lake_ChemistryPhysics_Raw.csv'), 
                          stringsAsFactors = TRUE)
NTL_LTER_Data$sampledate <- as.Date(NTL_LTER_Data$sampledate, format="%m/%d/%y")

# Checking if 'sampledate' is a date object
class(NTL_LTER_Data$sampledate)


#2 Here I build and set a custom ggplot theme
library(ggplot2)
custom_theme <- theme_minimal() +
  theme(
    text = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "blue"),
    panel.background = element_rect(fill = "lightgray"),
    panel.grid.major = element_line(color = "gray"),
    panel.grid.minor = element_blank()
  )
theme_set(custom_theme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not significantly change with depth.
Ha: Mean lake temperature recorded during July does significantly change with depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Here I Wrangle the NTL-LTER dataset
library(dplyr)
library(tidyr)

cleaned_data <- NTL_LTER_Data %>%
  filter(month(sampledate) == 7) %>%  
  select(lakename, year4, daynum, depth, temperature_C) %>%  
  drop_na()  

#5 Visualize the relationship between temperature and depth
ggplot(cleaned_data, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, color = "blue") +  
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  scale_y_continuous(limits = c(0, 35)) +  
  labs(
    title = "Relationship Between Lake Temperature and Depth in July",
    x = "Depth (m)",
    y = "Temperature (°C)"
  ) +
  theme_minimal() +  
  theme(
    text = element_text(size = 12, color = "black"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "blue"),
    panel.background = element_rect(fill = "lightgray"),
    panel.grid.major = element_line(color = "gray"),
    panel.grid.minor = element_blank()
  )

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
The scatter plot clearly shows a strong negative linear relationship between lake temperature and depth in July. As depth increases, temperature decreases. The linear regression line supports this trend, indicating that deeper water is significantly cooler than surface water. This observation aligns with typical lake stratification patterns, where warmer water tends to be concentrated near the surface.

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
linear_model <- lm(temperature_C ~ depth, data = cleaned_data)

# Display the results
summary(linear_model)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression analysis shows a significant negative relationship between lake temperature and depth, indicating that for every one-meter increase in depth, the temperature decreases by approximately (1.95 °C). The model explains (73.87%) of the variability in temperature, based on (9726 degrees of freedom). Both the intercept and depth coefficient are highly statistically significant (p < 2e-16), confirming that changes in depth have a meaningful impact on lake temperature.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 Here I created list of candidate models with different combinations of pred veriables
candidate_models <- list(
  model1 = lm(temperature_C ~ year4, data = cleaned_data),
  model2 = lm(temperature_C ~ daynum, data = cleaned_data),
  model3 = lm(temperature_C ~ depth, data = cleaned_data),
  model4 = lm(temperature_C ~ year4 + daynum, data = cleaned_data),
  model5 = lm(temperature_C ~ year4 + depth, data = cleaned_data),
  model6 = lm(temperature_C ~ daynum + depth, data = cleaned_data),
  model7 = lm(temperature_C ~ year4 + daynum + depth, data = cleaned_data)
)

# Calculate AIC for each model
AIC_values <- sapply(candidate_models, AIC)

# Find the model with the lowest AIC
best_model <- names(AIC_values)[which.min(AIC_values)]

# Display AIC values and the best model
AIC_values
best_model

#10 Then I Fit a multiple regression model with year4, daynum, and depth as predictors
multiple_regression_model <- 
  lm(temperature_C ~ year4 + daynum + depth, data = cleaned_data)

# Display the summary of the multiple regression model
summary(multiple_regression_model)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using year4, daynum, and depth as the final set of explanatory variables to predict lake temperature. This model explains approximately 74.1% of the observed variance in temperature, a slight improvement over the model with only depth, which explained around 73.9%. Adding variables enhances the model’s ability to explain temperature variation.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 First I fit the the ANOVA model
anova_model <- aov(temperature_C ~ lakename, data = cleaned_data)

# Summary of the ANOVA model
anova_summary <- summary(anova_model)

# Then i run the linear model (for comparison)
linear_model <- lm(temperature_C ~ lakename, data = cleaned_data)

# Display the summaries
anova_summary
summary(linear_model)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The ANOVA test shows a significant difference in mean temperature among the lakes in July (F(8, 9719) = 50, p < 2.2e-16). This indicates that at least one lake has a mean temperature that is significantly different from the others. The linear model confirms these results, with various lakes displaying significantly different temperatures compared to the reference lake. R^2 value of the linear model is 0.03953, suggesting that only about 3.95% of the variance in temperature is explained by the lake names.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14 Here I create a scatter plot of temperature by depth for each lake

ggplot(cleaned_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +  # 50% transparency
  geom_smooth(method = "lm", se = FALSE) +  # Linear model without confidence interval
  labs(title = "Lake Temperature by Depth in July",
       x = "Depth (m)",
       y = "Temperature (°C)",
       color = "Lake Name") +
  scale_y_continuous(limits = c(0, 35)) +  # Y-axis limits
  theme_minimal() +  # Use a minimal theme for clarity
  theme(text = element_text(size = 12),
        legend.position = "bottom")


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 performing Tukey's HSD test to determine which lakes have different means
library(stats)
tukey_test <- TukeyHSD(anova_model)

# Display the results
tukey_test

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the Tukey's HSD test results, Paul Lake is statistically similar to Peter Lake, with no significant difference in mean temperature (p = 0.2242). Additionally, Central Long Lake stands out as having a mean temperature that is statistically distinct from several other lakes, including East Long Lake, Hummingbird Lake, and Tuesday Lake, among others. Therefore, while Peter Lake shares a similar mean temperature with Paul Lake, Central Long Lake exhibits a significant difference compared to multiple lakes, indicating a unique thermal profile in July.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 

To compare the mean temperatures of Peter Lake and Paul Lake, we  can use a t-test. This test checks if there is a significant difference between the two lakes' average temperatures. If the p-value is less than 0.05, it indicates that the mean temperatures are significantly different.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18 First I Wrangle the data for Crampton Lake and Ward Lake
july_data <- cleaned_data %>%
  filter(lakename %in% 
           c("Crampton Lake", "Ward Lake"))

# Then I run the two-sample T-test
t_test_result <- t.test(temperature_C ~ lakename, data = july_data)

# Display the test results

print (t_test_result)

```

>Answer: The Welch Two Sample t-test comparing the July temperatures of Crampton Lake and Ward Lake showed a p-value of 0.2649. This means there is no significant difference in their mean temperatures; Crampton Lake's average is 15.35°C and Ward Lake's is 14.46°C. This result matches the earlier finding that no lakes have distinct mean temperatures.
